---
title: "Course Content"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The workshop is made up of 4 half-days. 

# Session 1

## Learning Contents

* R-studio 45'
* Use R to compute
  * Create an R script
  * Run lines or R script in total
  * Comparisons get a boolean out
  * Assign values to variables using `<-`
  * you can assign to itself
  * Vectorization
  * Use `ls()` to list variables
  * Use `rm()` to remove variables 
* libraries (5')
  * install libraries
  * load libraries
* Data structures (40')
  * tibbles / dataframes
  * pull out colums and do operations
  * data types `typeof()` command
* Factors (20')
  * Use for factors
  * tranform character into factor (use mpg data).
* tibbles
  * variable types
  * `head()` command
  * `summary()` commands
  *  calling a variable `$`operator
  
# Tidy Data
  * explain tidy data
  * Learn to create tidy data
  * gather and spread
  
# Manipulate Data
  * select()
    * negate arugument
  * filter()
  * arrange()
  * grpup_by()
  * summarise()
  * pipe
  

## Including Code

You can include R code in the document as follows:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
