---
title       : Introduction to R Workshop 
subtitle    : Session 2
author      : Muriel Lobier
job         : 
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : [mathjax, quiz, bootstrap]           # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---

<style>
pre {
  margin-top: -15px;
}
</style>

<style>
.large {
   font-size: 150%;
}
</style>

<style>
slide.smallcode pre {
  font-size: 16px;
}
</style>

---

## Session 2 - Main Goals

### Dataframes and Tibbles

###  Tidy data

### Data wrangling

---

## Dataframes and Tibbles

### Understand the avantages of tibbles

### Be able to extract information from tibbles

### Be able to manipulate factors
> Conver between string and factors

--- .class #id 

## Dataframes 

### Dataframes are the main data structures:
```{r}
iris
```

--- 

## Better dataframes: tibbles !!

### Load the `gapminder` library to access the gapminder dataset
```{r}
library(gapminder)
gapminder
```

--- 

## Better dataframes: Tibbles !!

### Why are tibbles better ???
```{r, echo=FALSE}
gapminder[1:5,]
```

```{r, echo=FALSE}
iris[1:5,]
```
---

## Better dataframes: Tibbles

### You can convert a data frame into a tibble
```{r}
as.tibble(iris)
```

--- 

## Main commands

### You can convert a data frame into a tibble
```{r}
names(gapminder)
```

--- 

## Main commands

```{r}
names(gapminder)
```
```{r}
dim(gapminder)
``` 

--- 

## Main commands

```{r}
names(gapminder)
```
```{r}
dim(gapminder)
``` 
```{r}
nrow(gapminder)
``` 

--- 

## Main commands

```{r}
names(gapminder)
```
```{r}
dim(gapminder)
``` 
```{r}
nrow(gapminder)
``` 
```{r}
ncol(gapminder)
``` 

---

## Specifying a variable

### Use the `$` operator

```{r}
summary(gapminder$gdpPercap)
```

---

## Categorical variables: factors

### Factors encode categorical variables

```{r}
gapminder
```


---

## Categorical variables: factors

### Factors have a finite number of levels

```{r}
nlevels(gapminder$continent)
```

### How many  countries are represented in the dataset??

---

## Categorical variables: factors

### Factor values can be extracted as a vector

```{r}
levels(gapminder$continent)
```

---

## Categorical variables: factors

### Factor values can be extracted as a vector

```{r}
levels(gapminder$continent)
```

### What are the last five countries in the dataset?

---

## Categorical variables: factors

### Factor values can be extracted as a vector

```{r}
levels(gapminder$continent)
```

### What are the last five countries in the dataset?
```{r}
tail(levels(gapminder$country))
```

---

## Categorical variables: factors

### How many observations do we have for each continent ?
```{r, eval=FALSE}
summary(gapminder$continent)
```

--- .smallcode

## Categorical variables: factors

### How many observations do we have for each continent ?
```{r}
summary(gapminder$continent)
```

---

## Categorical variables: factors

### How many observations do we have for each continent ?
```{r}
summary(gapminder$continent)
```


### How many observations do we have for each continent - other option?
```{r}
table(gapminder$continent)
```

--- 

## Categorical variables: factors

### How does R store factors "under the hood" ?
```{r, eval=FALSE}
str(gapminder$country)
```

---

## Converting string variables to factors

### Inspect the `mpg` data set - which variables should be factors ?

---

## Converting string variables to factors

### Inspect the `mpg` data set - which variables should be factors ?

```{r, echo=FALSE}
mpg[1:2,]
```

--- .smallcode

## Converting string variables to factors

### Inspect the `mpg` data set - which variables should be factors ?

```{r, echo=FALSE}
mpg[1:2,]
```

### Use the `as.factor()` command and  `<-`  to convert the `manufacturer` variable to  *factor*

--- .smallcode

## Converting string variables to factors

### Inspect the `mpg` data set - which variables should be factors ?

```{r, echo=FALSE}
mpg[1:2,]
```

### Use the `as.factor()` command and  `<-`  to convert the `manufacturer` variable to  *factor*

```{r}
mpg$manufacturer<-as.factor(mpg$manufacturer)
```
```{r, echo=FALSE}
mpg[1:2,]
```

--- .smallcode

## Converting string variables to factors

### Convert the `manufacturer` variable back to character (`as.character()`)

```{r, echo=FALSE}
mpg[1:2,]
```

--- .smallcode

## Converting string variables to factors

### Convert the `manufacturer` variable back to character (`as.character()`)

```{r}
mpg$manufacturer<-as.character(mpg$manufacturer)
```
```{r, echo=FALSE}
mpg[1:2,]
```

---

```{r, eval=FALSE}
str(gapminder$country)
```


---

## Categorical variables: factors

### How does R store factors "under the hood" ?
```{r, eval=FALSE}
str(gapminder$country)
```



### How many observations do we have for each continent - other option?
```{r}
table(gapminder$continent)
```


---

--- &twocol

## Subsetting Lists 

*** =left

```{r}
a <- list(a = 1:3, b = "a string", 
          c = pi, d = list(-1, -5))
str(a)
```


*** =right

<img class=center
  src=assets/img/session_1_r_list2.png height=100px>


*** =fullwidth

<br>
  Source: **H. Wickham** - [*R* for data science](http://r4ds.had.co.nz/vectors.html) - [CC license](https://creativecommons.org/licenses/by-nc-nd/3.0/us/)


