---
title       : Introduction to R Workshop 
subtitle    : Session 2
author      : Muriel Lobier
job         : 
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : [mathjax, quiz, bootstrap]           # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---

<style>
pre {
  margin-top: -15px;
}
</style>

<style>
.large {
   font-size: 150%;
}
</style>

<style>
slide.smallcode pre {
  font-size: 16px;
}
</style>

---

## Session 2 - Main Goals

### Dataframes and Tibbles

###  Tidy data

### Data wrangling

---

## Dataframes and Tibbles

### Understand the avantages of tibbles

### Be able to extract information from tibbles

### Be able to manipulate factors


--- .class #id 

## Dataframes 

### Dataframes are the main data structures:
```{r}
rm(list=ls())
iris
```

--- 

## Better dataframes: tibbles !!

### Load the `gapminder` library to access the gapminder dataset
```{r}
library(gapminder)
gapminder
```

--- 

## Better dataframes: Tibbles !!

### Why are tibbles better ???
```{r, echo=FALSE}
gapminder[1:5,]
```

```{r, echo=FALSE}
iris[1:5,]
```
---

## Better dataframes: Tibbles

### You can convert a data frame into a tibble
```{r}
as.tibble(iris)
```

--- 

## Main commands

### You can convert a data frame into a tibble
```{r}
names(gapminder)
```

--- 

## Main commands

```{r}
names(gapminder)
```
```{r}
dim(gapminder)
``` 

--- 

## Main commands

```{r}
names(gapminder)
```
```{r}
dim(gapminder)
``` 
```{r}
nrow(gapminder)
``` 

--- 

## Main commands

```{r}
names(gapminder)
```
```{r}
dim(gapminder)
``` 
```{r}
nrow(gapminder)
``` 
```{r}
ncol(gapminder)
``` 

---

## Specifying a variable

### Use the `$` operator

```{r}
summary(gapminder$gdpPercap)
```

---

## Categorical variables: factors

### Factors encode categorical variables

```{r}
gapminder
```


---

## Categorical variables: factors

### Factors have a finite number of levels

```{r}
nlevels(gapminder$continent)
```

### How many  countries are represented in the dataset??

---

## Categorical variables: factors

### Factor values can be extracted as a vector

```{r}
levels(gapminder$continent)
```

---

## Categorical variables: factors

### Factor values can be extracted as a vector

```{r}
levels(gapminder$continent)
```

### What are the last five countries in the dataset?

---

## Categorical variables: factors

### Factor values can be extracted as a vector

```{r}
levels(gapminder$continent)
```

### What are the last five countries in the dataset?
```{r}
tail(levels(gapminder$country))
```

---

## Categorical variables: factors

### How many observations do we have for each continent ?
```{r, eval=FALSE}
summary(gapminder$continent)
```

--- .smallcode

## Categorical variables: factors

### How many observations do we have for each continent ?
```{r}
summary(gapminder$continent)
```

---

## Categorical variables: factors

### How many observations do we have for each continent ?
```{r}
summary(gapminder$continent)
```


### How many observations do we have for each continent - other option?
```{r}
table(gapminder$continent)
```

--- 

## Categorical variables: factors

### How does R store factors "under the hood" ?
```{r, eval=FALSE}
str(gapminder$country)
```

---

## Converting string variables to factors

### Inspect the `mpg` data set - which variables should be factors ?

---

## Converting string variables to factors

### Inspect the `mpg` data set - which variables should be factors ?

```{r, echo=FALSE}
mpg[1:2,]
```

--- .smallcode

## Converting string variables to factors

### Inspect the `mpg` data set - which variables should be factors ?

```{r, echo=FALSE}
mpg[1:2,]
```

### Use the `as.factor()` command and  `<-`  to convert the `manufacturer` variable to  *factor*

--- .smallcode

## Converting string variables to factors

### Inspect the `mpg` data set - which variables should be factors ?

```{r, echo=FALSE}
mpg[1:2,]
```

### Use the `as.factor()` command and  `<-`  to convert the `manufacturer` variable to  *factor*

```{r}
mpg$manufacturer<-as.factor(mpg$manufacturer)
```
```{r, echo=FALSE}
mpg[1:2,]
```

--- .smallcode

## Converting string variables to factors

### Convert the `manufacturer` variable back to character (`as.character()`)

```{r, echo=FALSE}
mpg[1:2,]
```

--- .smallcode

## Converting string variables to factors

### Convert the `manufacturer` variable back to character (`as.character()`)

```{r}
mpg$manufacturer<-as.character(mpg$manufacturer)
```
```{r, echo=FALSE}
mpg[1:2,]
```

---

## Categorical variables: factors

### How does R store factors "under the hood" ?
```{r, eval=FALSE}
str(gapminder$country)
```

---

## Categorical variables: factors

### How does R store factors "under the hood" ?

```{r}
str(gapminder$country)
```


---

## Tidy data

### Be able to identify if a dataset is tidy

### Be able to use tidyverse functions to tidy a dataset

  * `gather()`
  * `spread()`
  
---

## Tidy data

### What is tidy data? 



---

--- &twocol_l

## Tidy data

### Run the following line of code to load 4 datasets
```{r}
load(file="data/is_my_data_tidy.RData")
```


*** =left

gapminder_1

gapminder_2

*** =right

gapminder_3

gapminder_4_1 and _2


--- .smallcode &twocol_l

## Tidy data

### Which one(s) of these datasets is(are) tidy?

*** =left

gapminder_1
```{r, echo=FALSE}
gapminder_1[1:4,]
```

gapminder_3
```{r, echo=FALSE}
gapminder_3[1:4,]
```

*** =right

gapminder_2
```{r, echo=FALSE}
gapminder_2[1:4,]
```

gapminder_4_1 and _2

```{r, echo=FALSE}
gapminder_4_1[1,]
```

```{r, echo=FALSE}
gapminder_4_2[1,]
```

---

## Tidy data
### gapminder_1
```{r, echo=FALSE}
gapminder_1
```

---

## Tidy data
### gapminder_1
```{r, echo=FALSE}
gapminder_1
```

**NOT TIDY**


---.smallcode


## Tidy data
### gapminder_2
```{r, echo=FALSE}
gapminder_2
```

--- .smallcode

## Tidy data
### gapminder_2
```{r, echo=FALSE}
gapminder_2
```

**NOT TIDY**

---

## Tidy data
### gapminder_3
```{r, echo=FALSE}
gapminder_3
```

---

## Tidy data
### gapminder_3
```{r, echo=FALSE}
gapminder_3
```

**TIDY**

---.smallcode

## Tidy data
### gapminder_4_1 and _4_2
```{r, echo=FALSE}
gapminder_4_1[1:4,]
```
```{r, echo=FALSE}
gapminder_4_2[1:4,]
```

---.smallcode

## Tidy data
### gapminder_4_1 and _4_2
```{r, echo=FALSE}
gapminder_4_1[1:4,]
```
```{r, echo=FALSE}
gapminder_4_2[1:4,]
```

**NOT TIDY**


---.smallcode

## Making data tidy - spread()

### What do we need to do to go from here:
```{r, echo=FALSE}
gapminder_1[1:4,]
```
to 

```{r, echo=FALSE}
gapminder_3[1:4,]
```

---

## Making data tidy - spread()

### `spread(data, key, value, fill=NA, ...)`

---.smallcode

## Making data tidy - spread()

### `spread(data, key = ? , value = ?)`

```{r, echo=FALSE}
my_data<-tibble(subject=c("S001", "S001", "S002", "S002"), measure_type=c("RT", "hit_rate","RT", "hit_rate"),  result=c(1.45, 0.71, 1.2,0.56))
```
```{r, echo=FALSE}
my_data
```

---.smallcode

## Making data tidy - spread()

### `spread(data, key = ? , value = ?)`

```{r, echo=FALSE}
my_data
```
```{r}
spread(data=my_data, key=measure_type, value = result)
```



---.smallcode

## Making data tidy - spread()

### Use the `spread()` function to go from:
```{r, echo=FALSE}
gapminder_1[1:4,]
```
to 

```{r, echo=FALSE}
gapminder_3[1:4,]
```

---.smallcode

## Making data tidy - spread()

### Use the `spread()` function:

```{r, echo=FALSE}
gapminder_1[1:4,]
```

```{r, eval=FALSE}
spread(gapminder_1, key = variable_name, value = variable_value)
```


---.smallcode

## Making data tidy - spread()

### Write the correct `spread()` function:

```{r, echo=FALSE}
gapminder_1[1:4,]
```

```{r}
spread(gapminder_1, key = variable_name, value = variable_value)
```


---.smallcode

## Making data tidy 

### How do we get from :

```{r, echo=FALSE}
gapminder_2[1:4,]
```
 to
 
```{r, echo=FALSE}
gapminder_3[1:4,]
```

---.smallcode

## Making data tidy - gather()

### Step 1: `gather()`

```{r, echo=FALSE}
gapminder_2[1:3,]
```
 to
 
```{r, echo=FALSE}
gather(gapminder_2, key=variable_type, value = variable_value, -country)[1:4,]
```

---.smallcode

## Making data tidy - separate()

### Step 2: `separate()`

```{r, echo=FALSE}
gapminder_2[1:3,]
```
 to
 
```{r, echo=FALSE}
gather(gapminder_2, key=variable_type, value = variable_value, -country) %>%
  separate(col = variable_type, into = c("var_type", "year")) %>% 
  head()
  
```

---.smallcode

## Making data tidy - gather()  again

### Step 3: `gather()`

```{r, echo=FALSE}
gather(gapminder_2, key=variable_type, value = variable_value, -country) %>%
  separate(col = variable_type, into = c("var_type", "year")) %>% 
  head()
```
 to
 
```{r, echo=FALSE}
gather(gapminder_2, key=variable_type, value = variable_value, -country) %>%
  separate(col = variable_type, into = c("var_type", "year")) %>% 
  spread(key=var_type, value=variable_value) %>%
  head()
  
```

---.smallcode

## Making data tidy - gather()

### `gather(data = , key = , value = )`

```{r, echo=FALSE}
tibble(subject=c("S001", "S001", "S001", "S001"), measure_type=c("RT.session1", "RT.session2","hitrate.session1", "hitrate.session2"),  result=c(1.45, 1.23, 0.74, 0.85)) %>%
  spread(key=measure_type, value=result) -> my_data
```
```{r, echo=FALSE}
my_data
```

---.smallcode

## Making data tidy - gather()

### `gather(data = , key = , value = )`

```{r, echo=FALSE}
my_data
```
```{r, eval=FALSE}
my_data_2 <- gather(data = my_data, key = variable, value = result, 
                    hitrate.session1, hitrate.session2, 
                    RT.session1, RT.session2)
```


---.smallcode

## Making data tidy - gather()

### `gather(data = , key = , value = )`

```{r, echo=FALSE}
my_data
```
```{r}
( my_data_2 <- gather(data = my_data, key = variable, value = result, hitrate.session1, 
                      hitrate.session2, RT.session1, RT.session2) )
```

---.smallcode

## Making data tidy - gather()

### `gather(data = , key = , value = )`

```{r, echo=FALSE}
my_data
```
```{r}
( my_data_2 <- gather(data = my_data, key = variable, value = result, hitrate.session1:RT.session2) )
```


---.smallcode

## Making data tidy - gather()

### `gather(data = , key = , value = )`

```{r, echo=FALSE}
my_data
```
```{r}
( my_data_2 <- gather(data = my_data, key = variable, value = result,  -subject) )
```


---.smallcode

## Making data tidy - gather()

### `gather(data = , key = , value = )`

```{r, echo=FALSE}
my_data
```
```{r}
gather(data = my_data, key = variable, value = result) 
```



---.smallcode

## Making data tidy - separate()

### `separate(data = , col = , into = , sep = , ...)`

```{r, echo=FALSE}
my_data_2
```
```{r}
( my_data_3 <- separate(data = my_data_2 , col = variable, into =  c("RT","session")) )
```

---.smallcode

## Making data tidy - spread again()

### `spread(data = , key = , value = )`

```{r, echo=FALSE}
my_data_3
```

---.smallcode

## Making data tidy - spread again()

### `spread(data = , key = , value = )`

```{r, echo=FALSE}
my_data_3
```
```{r}
( my_data_final <- spread(my_data_3, key=RT, value = result) )
```


---.smallcode

## Making data tidy - piping `%>%`

### Full tidying code : 

```{r, eval=FALSE}
my_data_2 <- gather(data = my_data, key = variable_type, value = result, -subject)
my_data_3 <- separate(data = my_data_2, col = variable_type, into = c('var_type', "session"))
my_data_final <- spread(data = my_data_3, key = var_type, value = result)
```


---.smallcode

## Making data tidy - piping `%>%`

### Full tidying code : 

```{r, eval = FALSE}
my_data_2 <- gather(data = my_data, key = variable_type, value = result, -subject)
my_data_3 <- separate(data = my_data_2, col = variable_type, into = c('var_type', "session"))
my_data_final <- spread(data = my_data_3, key = var_type, value = result)
```

### Piped tidying code : 

```{r}
my_data %>% 
  gather(key = variable_type, value = result, -subject) %>%
  separate(col = variable_type, into = c('var_type', "session")) %>%
  spread(key = var_type, value = result)
```


---.smallcode

## Making data tidy - piping `%>%`

### Full tidying code : 

```{r, eval = FALSE}
my_data_2 <- gather(data = my_data, key = variable_type, value = result, -subject)
my_data_3 <- separate(data = my_data_2, col = variable_type, into = c('var_type', "session"))
my_data_final <- spread(data = my_data_3, key = var_type, value = result)
```

### Piped tidying code : 

```{r}
my_data %>% 
  gather(key = variable_type, value = result, -subject) %>%
  separate(col = variable_type, into = c('var_type', "session")) %>%
  spread(key = var_type, value = result) -> my_data_final
```

---.smallcode

## Making data tidy - your turn !!
```{r, echo=FALSE}
gapminder_2[1:4,]
```
 to
 
```{r, echo=FALSE}
gapminder_3[1:4,]
```

---.smallcode

## Making data tidy - a solution
```{r, echo=FALSE}
gapminder_2[1:2,]
```
 to
 
```{r}
gapminder_2 %>%
  gather(key= variable_type, value=variable_value, -country) %>%
  separate(col=variable_type, into=c("var_type", "year")) %>%
  spread(key=var_type, value=variable_value)
```

---.smallcode
### Making data tidy - joining tibbles

```{r, echo=FALSE}
gapminder_4_1[1:2,]
```
```{r, echo=FALSE}
gapminder_4_2[1:2,]
```

---.smallcode

### Making data tidy - joining tibbles

```{r, echo=FALSE}
tibble(subject=c("S001", "S001", "S002", "S002"), 
       session=c("session1", "session2","session1", "session2"),  RT=c(1.45, 1.23, 1.74, 1.85))
```
```{r, echo=FALSE}
tibble(subject=c("S001", "S001", "S002", "S002"), 
       session=c("session1", "session2","session1", "session2"),  hitrate=c(.65, .73, .71, .68))
```


---.smallcode
### Making data tidy - joining datasets

```{r, echo=FALSE}
gapminder_4_1 %>% 
  gather(key=year, value=lifeExp,-country) %>%
  head()
```
```{r, echo=FALSE}
gapminder_4_2 %>% 
  gather(key=year, value=gdpPercap, -country) %>%
  head()
```


--- 

## Tidy data Joining datasets

<img class=center
  src=assets/img/session_2_join-setup.png width=300px>


--- .smallcode &twocol_l

## Tidy data Joining datasets

*** =left
### Inner join 

<img class=center
  src=assets/img/session_2_join-inner.png width=300px>

*** =right
### Outer join
<img class=center
  src=assets/img/session_2_join-outer.png width=300px>

*** =fullwidth

<br>
  Source: **H. Wickham** - [*R* for data science](http://r4ds.had.co.nz/vectors.html) - [CC license](https://creativecommons.org/licenses/by-nc-nd/3.0/us/)

---.smallcode

### Making data tidy - joining datasets
```{r}
gapminder_4_1 %>% 
  gather(key=year, value=lifeExp,-country) -> gapminder_4_1
```

```{r, echo=FALSE}
gapminder_4_1 %>% 
  gather(key=year, value=lifeExp,-country) %>%
  head()  
```


```{r}
gapminder_4_2 %>% 
  gather(key=year, value=gdpPercap,-country) -> gapminder_4_2
```


```{r, echo=FALSE}
gapminder_4_2 %>% 
  gather(key=year, value=gdpPercap, -country) %>%
  head()
```



```{r}
inner_join(gapminder_4_1, gapminder_4_2, by=c("country", "year"))
```


