---
title       : Introduction to R for reproducible data visualization and analysis 
subtitle    : Session 3
author      : Muriel Lobier
job         : 
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : [mathjax, quiz, bootstrap]           # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---

<style>
pre {
  margin-top: -15px;
}
</style>

<style>
.large {
   font-size: 150%;
}
</style>

<style>
slide.smallcode pre {
  font-size: 16px;
}
</style>

---

## Session 3 - Using R to run reproducible data analysis

### Project structure

### Importing data

### Visualizing data

### Reporting results

### Making informative plots with ggplot

---

## Open Data from a Posner experiment

### Paper [HERE](https://link.springer.com/article/10.3758%2Fs13428-017-0935-1) 

### Data [HERE](https://osf.io/cwzds/)


<img class=center
  src=assets/img/session_3_paper_header.png width=800px>


---

## Step 1: create a project folder structure

Project directories should be organized consistently and methodically. My preferred structure would be:

### project_name/ (root directory)
* name_of_analysis.Rmd (Rnotebook with main analysis code)
* R/ (project-specific R function files)
* data/ (original data - READ ONLY)
* figure/ (figures automatically saved when running code)
* figure_output (figures for use elsewhere)
* output/ (processed or simulated data, other non-final output )

**Navigate to R_Workshop_Exercises**


---

## Step 2: Open our R Notebook for our analysis

 


---.smallcode

## Importing data

### Use `read_csv()` to import the data

```{r, eval =  FALSE}
raw_data<-read_csv("data/study3_allsubjects_Posner.csv")
```

---.smallcode

## Importing data

### Use `read_csv()` to import the data

```{r}
raw_data<-read_csv("data/study3_allsubjects_Posner.csv")
```


---.smallcode

## Importing data

### Use `read_csv()` to import the data

Modify the code to not assign the first row to column names

```{r}
raw_data<-read_csv("data/study3_allsubjects_Posner.csv", col_names = FALSE)
```

---.smallcode

## Preprocess data: first glimpse of the dataset

### Quick overview: variable number and type, number of rows ....


---.smallcode

## Preprocess data: first glimpse of the dataset

### Quick overview: variable number and type, number of rows ....


```{r}
raw_data
```

---.smallcode

## Preprocess data: assign column names

### Use the `names()` command to assign names to variables
1. block
2. trial_number
3. SOA (delay between cue onset and target) in seconds
4. condition (0 = valid,  2=invalid)
5. response  (correct (1) or incorrect (0) )
6. RT: Reaction time (seconds)
7. Participant ID

---.smallcode

## Preprocess data: assign column names

### Use the `names()` command to assign names to variables
1. block
2. trial_number
3. SOA (delay between cue onset and target) in seconds
4. condition (0 = valid,  2=invalid)
5. response  (correct (1) or incorrect (0) )
6. RT: Reaction time (seconds)
7. Participant ID

```{r}
names(raw_data)<-c("block", "trial_number", "SOA", "condition", "response", "RT", "part_nb", "session_nb" )
```

---.smallcode

## Preprocess data: Modify variables ?


---.smallcode

## Preprocess data: Modify variables ?
```{r}
glimpse(raw_data)
```


---.smallcode

## Preprocess data: Modify variables ?
```{r}
summary(raw_data)
```

---.smallcode

## Preprocess data: Modify variables ?

Convert block and session to factors
```{r, eval = FALSE}
raw_data$block<- as.factor()
```

---.smallcode

## Preprocess data: Modify variables ?

Convert block and session to factors
```{r}
raw_data$block<- as.factor(raw_data$block)
raw_data$session_nb<- as.factor(raw_data$session_nb)
```

---.smallcode

## Preprocess data: Modify variables ?

Use `mutate()` and `if_else` to recode condition to descriptive values
```{r}
raw_data %>% 
  mutate(condition_f = if_else(condition == 0, "valid", "invalid")) -> raw_data
```


---.smallcode

## Check that participants were 'behaving well'

Use `group_by()` and `summarise()`  to inspect individual hit-rates for each session.




---.smallcode

## `group_by()` and `summarise()`

```{r, eval = FALSE}
data %>%
  group_by(session) %>%
  summarise(count = n(), mean = mean(RT))
```

<img class=center
  src=assets/img/session_3_group_by.png width=800px>

---.smallcode

## Check that participants were 'behaving well'

Use `group_by()` and `summarise()`  to inspect individual hit-rates for each session. 

```{r, eval=FALSE}
raw_data %>%
  group_by() %>%
  summarise()
```


---.smallcode

## Check that participants were 'behaving well'

Use `group_by()` and `summarise()` to inspect individual hit-rates for each session. 

```{r}
raw_data %>%
  group_by(part_nb, session_nb) %>%
  summarise(hit_rate = mean(response))
```

---.smallcode

## Check that participants were 'behaving well'

Use `group_by()` and `summarise()`  to inspect individual hit-rates for each session. 

A picture is worth a thousand tables, fill in the code to produce this plot:

```{r, fig.width=15,fig.height=10, fig.align="center",out.width="500px", echo = FALSE, warning=FALSE}
raw_data %>%
  group_by(part_nb, session_nb) %>%
  summarise(hit_rate = mean(response)) %>%
  ggplot(aes(x = session_nb, y = hit_rate)) + 
  geom_dotplot(binaxis='y', stackdir = "center")
```

---.smallcode

## Check that participants were 'behaving well'

Use `group_by()` and `summarise()`  to inspect individual hit-rates for each session. 

A picture is worth a thousand tables

```{r, fig.width=15,fig.height=10, fig.align="center",out.width="400px", message= FALSE}
raw_data %>%
  group_by(part_nb, session_nb) %>%
  summarise(hit_rate = mean(response)) %>%
  ggplot(aes(x = session_nb, y = hit_rate)) + 
  geom_dotplot(binaxis='y', stackdir = "center")
```



---.smallcode

## Explore individual data for possible outliers

* Write the ggplot code to produce this plot:

```{r, fig.width=15,fig.height=10, fig.align="center",out.width="700px", echo = FALSE}
ggplot(raw_data, aes(y = RT, x = condition_f) ) + 
  geom_boxplot(na.rm = TRUE) +
  facet_wrap(~part_nb)
```

---.smallcode

## Explore individual data for possible outliers

```{r, fig.width=15,fig.height=10, fig.align="center",out.width="600px"}
ggplot(raw_data, aes(y = RT, x = condition_f) ) + 
  geom_boxplot(na.rm = TRUE) +
  facet_wrap(~part_nb)
```

---.smallcode

## Explore individual data for possible outliers

### Use the filter funtion to remove the outlier data point of subject 23


```{r, eval = FALSE}
raw_data %>% 
  filter() -> raw_data
```

---.smallcode

## Use the filter funtion to remove the outlier

```{r}
raw_data %>% 
  filter(RT < 5) -> raw_data
```

---.smallcode

## Explore individual data for possible outliers

### What additional criteria could /should we apply ??

```{r, fig.width=15,fig.height=10, fig.align="center",out.width="680px", echo = FALSE}
ggplot(raw_data, aes(y = RT, x = condition_f) ) + 
  geom_boxplot(na.rm = TRUE) +
  facet_wrap(~part_nb)
```

---.smallcode

## Explore individual data for possible outliers

### Criteria applied by the original study:
* RTs less than 100 ms (no one is that fast)
* RTs above median RT +  3 times the median absolute deviation

---.smallcode

## Explore individual data for possible outliers

### Criteria applied by the original study:

* RTs less than 100 ms (Only The Flash is that fast)

### Use `summarise`,  `n()` and `group_by` to count the `nb_outliers` for each participant.

```{r, eval = FALSE}
raw_data %>%
  filter( ) %>%
  group_by() %>%
  summarise()
```


---.smallcode

## Explore individual data for possible outliers

### Criteria applied by the original study:

* RTs less than 100 ms (Only The Flash is that fast)

### Use `summarise`,  `n()` and `group_by` to count the `nb_outliers` for each participant.

```{r, eval = FALSE}
raw_data %>%
  filter( RT < .1) %>%
  group_by(part_nb) %>%
  summarise(nb_outliers = n())
```

---.smallcode

## Explore individual data for possible outliers

### Criteria applied by the original study:

* RTs less than 100 ms (Only The Flash is that fast)

### Now add ggplot code using  `geom_col()` to produce this plot:

```{r, fig.width=15,fig.height=7, fig.align="center",out.width="500px", eval=FALSE}
raw_data %>%
  filter( RT < .1) %>%
  group_by(part_nb) %>%
  summarise(nb_outliers = n()) %>%
  ggplot() +
  geom_col()
```

```{r, fig.width=15,fig.height=7, fig.align="center",out.width="500px", echo=FALSE}
raw_data %>%
  filter( RT < .1) %>%
  group_by(part_nb) %>%
  summarise(nb_outliers = n()) %>%
  ggplot(aes(x=part_nb, y=nb_outliers)) +
  geom_col()
```

---.smallcode

## Explore individual data for possible outliers

### Criteria applied by the original study:

* RTs below 100 ms (Only The Flash is that fast)

### Now add ggplot code using  `geom_col()` to produce this plot:


```{r, fig.width=15,fig.height=7, fig.align="center",out.width="500px"}
raw_data %>%
  filter( RT < .1) %>%
  group_by(part_nb) %>%
  summarise(nb_outliers = n()) %>%
  ggplot(aes(x=part_nb, y=nb_outliers)) +
  geom_col()
```


---.smallcode

## Explore individual data for possible outliers

### Criteria applied by the original study:
* RTs above median RT +  3 times the median absolute deviation 

### Use the `mad()` function to produce the same figure for this criteria 

```{r,  eval = FALSE}
raw_data %>%
  filter( ) %>%
  group_by(part_nb) %>%
  summarise(???) %>%
  ggplot(aes(x=part_nb, y=nb_outliers)) +
  geom_col()
```

---.smallcode

## Explore individual data for possible outliers

### Criteria applied by the original study:
* RTs above median RT +  3 times the median absolute deviation 

### Use the `mad()` function to produce the same figure for this criteria 


```{r, fig.width=15,fig.height=7, fig.align="center",out.width="500px"}
raw_data %>%
  group_by(part_nb) %>%
  filter( RT > median(RT) + 3*mad(RT)) %>%
  summarise(nb_outliers = n()) %>%
  ggplot(aes(x=part_nb, y=nb_outliers)) +
  geom_col()
```


---.smallcode

## Explore individual data for possible outliers

### Put it all together to remove outliers
* RTs below 100 ms (no one is that fast)
* RTs above median+ 3 times the median absolute deviation 

Fill in the code to remove all trials with outlier RTs

```{r, eval = FALSE}
raw_data %>%
  group_by(part_nb) %>%
  filter( ) -> preproc_data
  
```

---.smallcode

## Explore individual data for possible outliers

### Put it all together to remove outliers
* RTs over median+ 3 times the median absolute deviation 
* RTs less than 100 ms (only The Flash is that fast)

```{r}
raw_data %>%
  group_by(part_nb) %>%
  filter((RT > 0.1 & (RT < median(RT) + 3*mad(RT)))) -> preproc_data
  
```

---.smallcode

## Visualize preprocessed individual data

```{r, fig.width=15,fig.height=10, fig.align="center",out.width="680px", eval = FALSE}
ggplot( ) + 
  geom_boxplot() +
  facet_wrap()
```

```{r, fig.width=15,fig.height=10, fig.align="center",out.width="680px", echo = FALSE}
ggplot(preproc_data, aes(y = RT, x = condition_f) ) + 
  geom_boxplot(na.rm = TRUE) +
  facet_wrap(~part_nb)
```

---.smallcode

## Visualize preprocessed individual data

```{r, fig.width=15,fig.height=10, fig.align="center",out.width="680px"}
ggplot(preproc_data, aes(y = RT, x = condition_f) ) + 
  geom_boxplot(na.rm = TRUE) +
  facet_wrap(~part_nb)
```



---.smallcode

## Compute individual performance 

### Dependent/measured  variable: 
* RT 

### Independent/manipulated variables: 
* condition_f 
* session_nb

---.smallcode

## Compute individual performance 

### Dependent/measured  variable: 
* RT 

### Independent/manipulated variables: 
* condition_f 
* session_nb

### Compute individual mean RTs for each condition and session  
```{r, eval = FALSE}
preproc_data %>%
  group_by() %>%
  summarise() -> indiv_data
```

---.smallcode

## Evaluate individual performance 

### Dependent/measured  variable: 
* RT 

### Independent/manipulated variables: 
* condition_f 
* session_nb

### Compute individual mean RTs for each condition and session  
```{r}
preproc_data %>%
  group_by(part_nb, condition_f, session_nb) %>%
  summarise(RT_mean = mean(RT, na.rm=TRUE) , 
            RT_sd = sd(RT, na.rm=TRUE)) -> indiv_data
```

---.smallcode

## The classical plot 

```{r, fig.width=15,fig.height=10, fig.align="center",out.width="680px", echo = FALSE}
indiv_data %>%
  group_by(session_nb,condition_f) %>% 
  summarise(RT = mean(RT_mean), 
            se = (sd(RT_mean)/sqrt(n()))) %>%
  ggplot(aes(x = condition_f, y = RT, color = session_nb)) +
  geom_line(aes(group = session_nb), position = position_dodge(0.1)) +
  geom_linerange(aes(ymin= RT - se, ymax = RT + se), position = position_dodge(0.1)) +
  geom_point(position = position_dodge(0.1), size = 4) + 
  theme_classic() +
  theme(text = element_text(size = 20))
  
```


---.smallcode

## The classical plot - DON'T DO IT

```{r, fig.width=15,fig.height=10, fig.align="center",out.width="680px", echo = FALSE}
indiv_data %>%
  group_by(session_nb,condition_f) %>% 
  summarise(RT = mean(RT_mean), 
            se = (sd(RT_mean)/sqrt(n()))) %>%
  ggplot(aes(x = condition_f, y = RT, color = session_nb)) +
  geom_line(aes(group = session_nb), position = position_dodge(0.1)) +
  geom_linerange(aes(ymin= RT - se, ymax = RT + se), position = position_dodge(0.1)) +
  geom_point(position = position_dodge(0.1), size = 4) + 
  theme_classic() +
  theme(text = element_text(size = 20))
  
```


---

## Informative plotting of data

<img class=center
  src=assets/img/session_3_between_figures.png width=650px>


[Weissgerber TL, Milic NM, Winham SJ, Garovic VD (2015) Beyond Bar and Line Graphs: Time for a New Data Presentation Paradigm.](http://journals.plos.org/plosbiology/article?id=10.1371/journal.pbio.1002128)

---

## Informative plotting of data

<img class=center
  src=assets/img/session_3_group_comps.png width=450px>


[Rousselet, G. A., Foxe, J. J. and Bolam, J. P. (2016), A few simple steps to improve the description of group results in neuroscience. Eur J Neurosci, 44: 2647-2651.](http://onlinelibrary.wiley.com/doi/10.1111/ejn.13400/full)

---.smallcode &twocol_l

## Plotting a 2 x 2 within participant design: your choice !

*** =left

### Option 1: visualize the direction of individual differences

```{r, fig.width=15,fig.height=10, fig.align="center",out.width="400px", echo = FALSE, message=FALSE}
ggplot(indiv_data, aes(x = condition_f, y = RT_mean) ) + 
  geom_boxplot(na.rm = TRUE, fill = "light blue", outlier.color = "red", color = "blue") +
  geom_point(shape = 1, size = 2, color = "blue") +
  geom_line(aes(group = part_nb),linetype = 2 ) +
  theme_bw() + 
  facet_wrap(~session_nb, labeller = label_both) +
  theme(text = element_text(size = 24))
```

*** =right

### Option 2: Plot within-subject RT differences between invalid and valid trials

```{r, fig.width=15,fig.height=10, fig.align="center",out.width="450px", echo = FALSE, message = FALSE}
indiv_data %>% select(-RT_sd) %>%
  spread(key = condition_f, value = RT_mean) %>%
  mutate(rt_diff = invalid-valid) %>%
  ggplot(aes(x=session_nb, y=rt_diff)) +
    geom_hline(yintercept = 0, color = "orange", size = 2) +
    geom_violin(fill = "red", alpha = 0.1, color = NA) + 
    geom_boxplot(color="red", outlier.shape = NA, width = .3, alpha = 0.5, size = 1) + 
    geom_dotplot(binaxis ="y", stackdir="center", fill="red", alpha = 0.3) +
    theme_bw() +
    theme(text = element_text(size = 30)) +
    ylab("RT_invalid - RT_valid") 
  
```

---.smallcode
## Exporting your plot

Use `ggsave()` to export your plot to a variety of formats:
* "png", "eps", "ps", "pdf", "jpeg", "tiff", "png", "bmp", "svg"

* Arguments
```{r, eval= FALSE}
ggsave(filename, plot = last_plot(), device = NULL, path = NULL,
  scale = 1, width = NA, height = NA, units = c("in", "cm", "mm"),
  dpi = 300, limitsize = TRUE, ...)
```

* Examples
```{r, eval = FALSE}
ggsave("figure_ouput/myfig.pdf")
ggsave("figure_ouput/myfig.png")
ggsave("figure_ouput/myfig.pdf", width = 8.7, height = 10, units = "cm")
```




---.smallcode

## Next analysis step : Inferential statistics !

### Our hypotheses:
* RTs are faster for valid than invalid trials
* This difference in RTs varies across sessions

---.smallcode

## Next analysis step : Inferential statistics !

### Our hypotheses:
* RTs are faster for valid than invalid trials
* This difference in RTs varies across sessions

### One possible statistical test: the 2 X 2 Repeated Measures ANOVA
* Factor 1: condition_f 
* Factor 2: session_nb

---.smallcode

## Next analysis step : Inferential statistics !

### Our hypotheses:
* RTs are faster for valid than invalid trials
* This difference in RTs varies across sessions

### One possible statistical test: the 2 X 2 Repeated Measures ANOVA
* Factor 1: condition_f 
* Factor 2: session_nb

--- .smallcode

### A package for ANOVAs made easy: ez

```{r, warning = FALSE}
library(ez)
```

```{r, eval = FALSE}
anova_model <- ezANOVA(data = indiv_data,
                       dv = RT_mean,
                       wid = part_nb,
                       within = .(condition_f, session_nb))
```

--- .smallcode

### Reproducible ANOVA analysis

```{r}
anova_model <- ezANOVA(data = indiv_data,
                       dv = RT_mean,
                       wid = part_nb,
                       within = .(condition_f, session_nb))
```

```{r}
anova_model
```

Add the ANOVA analysis to your analysis file as well as a paragraph explaining the results.

--- .smallcode

### Reproducible data analysis

Your reproducible data analysis file is now ready and complete. 

Press the *Knit* button to process your file into an html document.

Well done !!!



---

## Informative plotting of data - independent groups

<img class=center
  src=assets/img/session_3_group_comps.png width=450px>


[Rousselet, G. A., Foxe, J. J. and Bolam, J. P. (2016), A few simple steps to improve the description of group results in neuroscience. Eur J Neurosci, 44: 2647-2651.](http://onlinelibrary.wiley.com/doi/10.1111/ejn.13400/full)

---
## Informative plotting of data - independent groups

Create a new .Rmd file in R/R_Workshop/Exercises

Use this .Rmd file and R chuncks to keep a trace of all you code !

First we will *modify* our data to simulate independent group data

---.smallcode

## Plotting options for fully independent groups
Use `rename()`and `mutate`  to change 
* condition_f to group : 
  * "valid" to "CTRL"
  * "invalid" to "DYS"
* session_nb to age
  * "1" to "young"
  * "2" to "old"

```{r, eval = FALSE}
indiv_data %>%
  ungroup() %>%
  select(-part_nb) %>%
  rename()%>% 
  mutate() -> indep_data
  
```


---.smallcode

## Plotting options for fully independent groups
Use `rename()`and `mutate`  to change 
* condition_f to group : 
  * "valid" to "CTRL"
  * "invalid" to "DYS"
* session_nb to age
  * "1" to "young"
  * "2" to "old"

```{r}
indiv_data %>%
  ungroup() %>%
  select(-part_nb) %>%
  rename(group = condition_f, age = session_nb)%>% 
  mutate(group = ifelse( group == "valid", "CTRL", "DYS"),
         age = ifelse(age == 1, "young" , "old")) -> indep_data
  
```


---.smallcode &twocol_l

## Plotting options for fully independent groups: your choice !

*** =left

### Option 1: boxplot and jittered individua data

```{r, fig.width=15,fig.height=10, fig.align="center",out.width="450px", echo = FALSE, message=FALSE}
ggplot(indep_data, aes(x = age, y = RT_mean, fill= group) ) + 
  geom_boxplot(na.rm = TRUE, width = 0.4, outlier.shape = NA, alpha = 0.7) +
   geom_point(pch = 21, position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.4), size = 2)+
  theme_bw() +
  ggtitle("Effect of reading ability and age on reaction times") +
  xlab("Session") + ylab("Mean reaction times") + 
  theme(text=element_text(size = 28)) +
   scale_fill_discrete(name="Group",
                          breaks=c("CTRL", "DYS"),
                         labels=c("Normal readers", "Dyslexic readers"))
```

*** =right

### Option 2: Pirate plots : yarrr !!!

* load the yarrr library
```{r}
library(yarrr)
```


```{r, fig.width=15,fig.height=10, fig.align="center",out.width="450px", echo = FALSE, message = FALSE}


pirateplot(formula = RT_mean ~ group + age,
           data = indep_data,
           theme = 1,
           pal = "xmen",
           point.pch = 16,
           gl.col = gray(.8),
           main = "Effect of cue validity and session on reaction times")
  
```


---.smallcode

## Plotting options for fully independent groups

Option 1: boxplot and jittered individua data

```{r, fig.width=15,fig.height=10, fig.align="center",out.width="450px"}
ggplot(indep_data, aes(x = age, y = RT_mean, fill= group) ) + 
  geom_boxplot(na.rm = TRUE, width = 0.4, outlier.shape = NA, alpha = 0.7) +
   geom_point(pch = 21, position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.4), size = 2)+
  theme_bw() +
  ggtitle("Effect of reading ability and age on reaction times") +
  xlab("Session") + ylab("Mean reaction times") + 
  theme(text=element_text(size = 28))
   scale_fill_discrete(name="Group",
                          breaks=c("CTRL", "DYS"),
                         labels=c("Normal readers", "Dyslexic readers"))
```

---.smallcode

## Plotting options for fully independent groups

Option 2: Pirate plots : yarrr !!!

```{r, fig.width=15,fig.height=10, fig.align="center",out.width="450px", message = FALSE}


pirateplot(formula = RT_mean ~ group + age,
           data = indep_data,
           theme = 1,
           pal = "xmen",
           point.pch = 16,
           gl.col = gray(.8),
           main = "Effect of cue validity and session on reaction times")
  
```


--- Create informative ERP plots

## Plotting ERP data

<img class=center
  src=assets/img/session_3_ERP_plots.png width=520px>


[Rousselet, G. A., Foxe, J. J. and Bolam, J. P. (2016), A few simple steps to improve the description of group results in neuroscience. Eur J Neurosci, 44: 2647-2651.](http://onlinelibrary.wiley.com/doi/10.1111/ejn.13400/full)


---.smallcode

## Create informative ERP plots

### Load the data from the ERP library

```{r}
library(ERP)
data(erpcz)
erp_data<-as.tibble(erpcz)
```

### Wrangle the data to get the following tibble

```{r}
erp_data %>%
  gather(key = x_time, value=amplitude, -Subject, -Instruction) %>%
  separate(col = x_time, into=c("x","time"), sep="X") %>%
  mutate(time = as.numeric(time)) ->erp_data
```


---.smallcode

## Create informative ERP plots

### Wrangle the data to get the following tibble

```{r, eval = FALSE}
erp_data %>%
  gather() %>%
  separate() %>%
  mutate() ->erp_data
```

```{r, echo = FALSE}
erp_data %>%
  gather(key = x_time, value=amplitude, -Subject, -Instruction) %>%
  separate(col = x_time, into=c("x","time"), sep="X") %>%
  mutate(time = as.numeric(time)) ->erp_data
erp_data
```

---.smallcode

## Create informative ERP plots

### Wrangle the data to get the following tibble

```{r, eval = FALSE}
erp_data %>%
  gather(key = x_time, value=amplitude, -Subject, -Instruction) %>%
  separate(col = x_time, into=c("x","time"), sep="X") %>%
  mutate(time = as.numeric(time)) -> erp_data
```

```{r}
erp_data
```


---.smallcode

## Create informative ERP plots

### Load the data from the ERP library

```{r}
library(ERP)
data(erpcz)
erp_data<-as.tibble(erpcz)
```

### Wrangle the data to get the following tibble

```{r}
erp_data %>%
  gather(key = x_time, value=amplitude, -Subject, -Instruction) %>%
  separate(col = x_time, into=c("x","time"), sep="X") %>%
  mutate(time = as.numeric(time)) ->erp_data
```

---.smallcode

## Create informative ERP plots

First create this non-informative plot of the group-averaged ERP

```{r, fig.width=15,fig.height=10, fig.align="center",out.width="680px", echo = FALSE}
erp_data %>% 
  group_by(Instruction, time) %>%
  summarise(Mean=mean(amplitude)) %>%
ggplot(aes(x=time, y = Mean, color= Instruction, fill = Instruction)) + 
  geom_line(size = 3) +
  theme_bw() + 
  theme(text = element_text(size = 28))
```

---.smallcode

## Create informative ERP plots

First create this non-informative plot of the group-averaged ERP

```{r, fig.width=15,fig.height=10, fig.align="center",out.width="450px"}
erp_data %>% 
  group_by(Instruction, time) %>%
  summarise(Mean=mean(amplitude)) %>%
ggplot(aes(x=time, y = Mean, color= Instruction, fill = Instruction)) + 
  geom_line(size = 3) +
  theme_bw()  + 
  theme(text = element_text(size = 28))
```

---.smallcode

## Create informative ERP plots

Now let's add the individual ERPs in the background

```{r, fig.width=15,fig.height=10, fig.align="center",out.width="680px", echo = FALSE}

erp_data %>% 
  group_by(Instruction, time) %>%
  summarise(Mean=mean(amplitude)) -> mean_erp_data

ggplot(erp_data, aes(x=time, y = amplitude, color= Instruction)) + 
  geom_line(aes(color= Instruction, group = interaction(Subject, Instruction)), alpha = 0.5, size =0.2) +
  geom_line(data = mean_erp_data, aes(y=Mean), size =2) +
  theme_bw()  + 
  theme(text = element_text(size = 28))
```

---.smallcode

## Create informative ERP plots

Now let's add the individual ERPs in the background

```{r, fig.width=15,fig.height=10, fig.align="center",out.width="480px"}

erp_data %>% 
  group_by(Instruction, time) %>%
  summarise(Mean=mean(amplitude)) -> mean_erp_data

ggplot(erp_data, aes(x=time, y = amplitude, color= Instruction)) + 
  geom_line(aes(color= Instruction, group = interaction(Subject, Instruction)), alpha = 0.5, size =0.2) +
  geom_line(data = mean_erp_data, aes(y=Mean), size =2) +
  theme_bw() + 
  theme(text = element_text(size = 28))
```


---.smallcode

## Create informative ERP plots - differences

Compute the difference between conditions

```{r, eval = FALSE}
erp_data %>% 
  spread(key = , value = ) %>%
  mutate() -> diff_erp_data
diff_erp_data
```

```{r, echo = FALSE}
erp_data %>% 
  spread(key = Instruction, value = amplitude) %>%
  mutate(amplitude_diff = TBR - TBF) -> diff_erp_data
diff_erp_data
```

---.smallcode

## Create informative ERP plots - differences

Compute the difference between conditions

```{r}
erp_data %>% 
  spread(key = Instruction, value = amplitude) %>%
  mutate(amplitude_diff = TBR - TBF) -> diff_erp_data
diff_erp_data
```

---.smallcode

## Create informative ERP plots - differences

Compute non-parametric confidence intervals using bootstrapping

* load the `rcompanion` library

```{r}
library(rcompanion)

```

---.smallcode

## Create informative ERP plots - differences

Compute non-parametric confidence intervals using bootstrapping

* Compute the ajusted bootstrapped percentile confidence intervals
```{r}
erp_boot_data <- groupwiseMean(amplitude_diff ~ time ,
              data   = diff_erp_data,
              conf   = 0.95,
              digits = 3,
              R      = 1000,
              boot        = TRUE,
              traditional = FALSE,
              normal      = FALSE,
              basic       = FALSE,
              percentile  = FALSE,
              bca         = TRUE)
head(erp_boot_data)
```


---.smallcode

## Create informative ERP plots

Create this plot using `geom_ribbon` to visualize the confidence intervals

```{r, fig.width=15,fig.height=10, fig.align="center",out.width="580px", echo = FALSE}
ggplot(erp_boot_data, aes(x=time, y = Mean)) + 
  geom_hline(yintercept = 0, color='red') +
  geom_ribbon(aes(ymin= Bca.lower, ymax= Bca.upper), alpha = 0.2) + 
  geom_line() +
  theme_bw()
```

---.smallcode

## Create informative ERP plots

Create this plot using `geom_ribbon` to visualize the confidence intervals

```{r, fig.width=15,fig.height=10, fig.align="center",out.width="480px"}
ggplot(erp_boot_data, aes(x=time, y = Mean)) + 
  geom_hline(yintercept = 0, color='red') +
  geom_ribbon(aes(ymin= Bca.lower, ymax= Bca.upper), alpha = 0.2) + 
  geom_line() +
  theme_bw()
```


---.smallcode

## Create informative ERP plots

Modify your plot to add individual differences in ERP response

```{r, fig.width=15,fig.height=10, fig.align="center",out.width="580px", echo = FALSE}

ggplot(erp_boot_data, aes(x=time, y = Mean)) + 
  geom_hline(yintercept = 0, color='red') +
  geom_line(data=diff_erp_data, aes(x=time, y=amplitude_diff, group = Subject ), color= "light gray", size =0.2) +
  geom_ribbon(aes(ymin= Bca.lower, ymax= Bca.upper), fill= NA, color = "blue",size = 1) + 
  geom_line(size = 2, color = "blue") +
  theme_bw()
```

---.smallcode

## Create informative ERP plots

Modify your plot to add individual differences in ERP response

```{r, fig.width=15,fig.height=10, fig.align="center",out.width="480px"}

ggplot(erp_boot_data, aes(x=time, y = Mean)) + 
  geom_hline(yintercept = 0, color='red') +
  geom_line(data=diff_erp_data, aes(x=time, y=amplitude_diff, group = Subject ), color= "light gray", size =0.2) +
  geom_ribbon(aes(ymin= Bca.lower, ymax= Bca.upper), fill= NA, color = "blue",size = 1) + 
  geom_line(size = 2, color = "blue") +
  theme_bw()
```

---.smallcode

## Create informative ERP plots

Time-course of individual differences with colors mapping amplitudes

```{r, fig.width=15,fig.height=10, fig.align="center",out.width="580px", echo = FALSE}

library(viridis)

ggplot(diff_erp_data, aes(x=time, y = Subject)) + 
  geom_raster(aes(fill = amplitude_diff)) +
  scale_x_continuous( limits = c(0,1000),expand = c(0, 0)) +
  scale_y_discrete( expand = c(0, 0)) +
  scale_fill_viridis() +
  theme_bw()
```

---.smallcode

## Create informative ERP plots

Time-course of individual differences with colors mapping amplitudes

```{r, fig.width=15,fig.height=10, fig.align="center",out.width="400px"}

library(viridis)

ggplot(diff_erp_data, aes(x=time, y = Subject)) + 
  geom_raster(aes(fill = amplitude_diff)) +
  scale_x_continuous( limits = c(0,1000),expand = c(0, 0)) +
  scale_y_discrete( expand = c(0, 0)) +
  scale_fill_viridis() +
  theme_bw()
```


---

## Session 3 - Using R to run reproducible data analysis

### Project structure

### Importing data

### Visualizing data

### Reporting results

### Making informative plots with ggplot


---

## Making informative plots with ggplot

### Be able to use *geoms*
* Use `geom_col()` to visualize counts
* Use `geom_line()` to link selected data points
* Use `geom_ribbon()` to visualize confidence intervals
* Use `geom_raster()` to make heatmaps


### Be able to modify a plot to improve visualization
* transform axes or color scales
* use `theme()` to modify text size
* Use `xlab()`, `ylab()`, `ggtitle()`

